Carrier phase and amplitude estimation for phase shift keying using pilots and data

### 2012-10-26 other/confnoncoherent/paper.tex
Added more of Barry's edits.

The paper is organised in the following way.  Section~\ref{sec:circ-symm-compl} describes properties of complex random variables that we need.  Section~\ref{sec:stat-prop-least} describes the statistical properties of the least squares estimators of carrier phase $\hat{\theta}$ and amplitude $\hat{\rho}$.  We show, under some assumptions about the distribution of the noise $w_1,\dots,w_L$, that $\langle\theta_0 - \hat{\theta}\rangle^2$ converges almost surely to zero and that $\sqrt{L}\langle\theta_0 - \hat{\theta}\rangle^2$ is asymptotically normally distributed as $L\rightarrow \infty$.  However, $\hat{\rho}$ is not a consistent estimator of the amplitude $\rho_0$.  The bias of $\hat{\rho}$ is small when the signal to noise ratio (SNR) is large, but the bias is significant when the SNR is small.  %In Section~\ref{sec:gaussian-noise-case} we consider the special case when the noise is Gaussian.  In this case, our formula for the asymptotic variance of can be simplified.  
Section~\ref{sec:simulations} presents the results of Monte-Carlo simulations.  These simulations agree with the derived asymptotic properties.

@@

Before describing the statistical properties of the least squares estimator, we first require some properties of complex valued random variables.  
A complex random variable $X$ is said to be \emph{circularly symmetric} if the distribution of its phase $\angle{X}$ is uniform on $[0,2\pi)$ and is independent of the distribution of its magnitude $\abs{X}$.  That is, if $Z \geq 0$ and $\Theta \in [0,2\pi)$ are real random variables such that $Ze^{j\Theta} = X$, then $\Theta$ is uniformly distributed on $[0,2\pi)$ and is independent of $Z$.  If the probability density function (pdf) of $Z$ is $f_Z(z)$, then the joint pdf of $\Theta$ and $Z$ is $f_{Z,\Theta}(z,\theta) = \frac{1}{2\pi}f_Z(z)$.
If $X$ is circularly symmetric, then for any real number $\phi$, the distribution of $X$ is the same as that of $e^{j\phi}X$.  %If $\expect\abs{X} = \expect Z$ is finite, then $X$ has zero mean because
% \begin{align*}
% \expect X &= \int_{0}^{2\pi} \int_{0}^\infty z e^{j\theta} f_{Z,\Theta}(z,\theta) dz d\theta \\
% &= \frac{1}{2\pi} \int_{0}^{2\pi} e^{j\theta} \int_{0}^\infty z f_Z(z) dz d\theta \\
% &= \frac{1}{2\pi}\expect Z \int_{0}^{2\pi} e^{j\theta} d\theta = 0.
% \end{align*}

We will have particular use of complex random variables of the form $1 + X$ where $X$ is circularly symmetric.  Let $R \geq 0$ and $\Phi \in [0,2\pi)$ be real random variables satisfying $R e^{j\Phi} = 1 + X$.  The joint distribution of $R$ and $\Phi$ can be shown to be
\[
f(r,\phi) = \frac{r f_Z(\sqrt{r^2 - 2r\cos\phi + 1})}{2\pi\sqrt{r^2 - 2r\cos\phi + 1}}.
\]

@@

\begin{theorem}\label{thm:normality} (Asymptotic normality)
Under the same conditions as Theorem~\ref{thm:consistency}, let $f(r,\phi)$ be the joint pdf of $R_1$ and $\Phi_1$, and let
\[
g(\phi) = \int_{0}^{\infty} r f(r,\phi) dr.
\]
Put $\hat{\lambda}_L = \sfracpart{\theta_0 - \hat{\theta}}$ and $\hat{m}_L = \hat{\rho} - \rho_0 G(0)$. %If $g(\phi)$ is continuous at $\phi = \tfrac{2\pi}{M}k+\tfrac{\pi}{M}$ for each $k = 0, 1, \dots M-1$, then 
Then the distribution of $(\sqrt{L}\hat{\lambda}_L, \sqrt{L}\hat{m}_L)$ converges to the bivariate normal with zero mean and covariance
\[
\left( \begin{array}{cc} 
H^{-2} A & 0 \\
0 & \rho_0^2 B
\end{array} \right)
\]
as $L \rightarrow \infty$, where
\[
H = G(0) -  2 \sin(\tfrac{\pi}{M}) \sum_{k = 0}^{M-1} g(\tfrac{2\pi}{M}k + \tfrac{\pi}{M}),
\]
\[
A = \expect R_1^2\sin^2\fracpart{\Phi_1}, \;\;\; B = \expect R_1^2 \cos^2\fracpart{\Phi_1} - G^2(0). 
\]
\end{theorem}

@@

A key assumption in Theorem~\ref{thm:consistency} is that $G(x)$ is uniquely maximised at $x = 0$ for $x \in [-\pi, \pi)$.  %This assumption asserts that $G(x) \leq G(0)$ for all $x \in [-\pi, \pi)$ and that if $\{x_i\}$ is a sequence of numbers from $[-\pi,\pi)$ such that $G(x_i) \rightarrow G(0)$ as $i \rightarrow \infty$ then $x_i \rightarrow 0$ as $i \rightarrow \infty$.  
Although we will not prove it here, this assumption is not only sufficient, but also necessary, for if $G(x)$ is uniquely maximised at some $x \neq 0$ then $\sfracpart{\theta_0 - \hat{\theta}}_\pi \rightarrow x$ almost surely as $L\rightarrow\infty$, while if $G(x)$ is not uniquely maximised then $\sfracpart{\theta_0 - \hat{\theta}}_\pi$ will not converge.  One can check that the assumption holds when $w_1$ is circularly symmetric and normally distributed.  %We will not attempt to further classify those distributions for which the assumption holds here.
%The next proposition describes a class of circularly symmetric distributions for which $G(x)$ is uniquely maximised at $x = 0$.  The zero mean complex Gaussian distribution with independent real and imaginary parts is in this class.  We omit the proof due to space constraints.

@@

The theorems place conditions on $\sfracpart{\hat{\theta} - \theta_0}$ rather than directly on $\hat{\theta} - \theta_0$.  This makes practical sense, and to see why let $s_i' = e^{j2\pi k/M}s_i$ be $M$-PSK symbols obtained by rotating $s_1, \dots, s_L$ by $e^{j2\pi k/M}$ for some integer $k$.  Then
\[
a s_i =  \rho e^{j\theta}s_i = \rho e^{j(\theta - 2\pi k/M)} s_i',
\]
and so, if $a = \hat{\rho}e^{j\hat{\theta}}$ minimises~\eqref{eq:hatadef}, then so does $\hat{\rho} e^{j(\hat{\theta} - 2\pi k/M)}$.  Thus, $\theta_0 - \hat{\theta}$ should be attributed the same error as $\theta_0 - \hat{\theta} - \tfrac{2\pi}{M}k$, that is, the error should be computed `modulo $\tfrac{2\pi}{M}$'.  It is for this reason that \emph{differential encoding} is often used with noncoherent $M$-PSK  detectors.

>>>>

The paper is organised in the following way.  Section~\ref{sec:circ-symm-compl} describes properties of complex random variables that we need.  Section~\ref{sec:stat-prop-least} describes the statistical properties of the least squares estimators of carrier phase $\hat{\theta}$ and amplitude $\hat{\rho}$.  We show, under some assumptions about the distribution of the noise $w_1,\dots,w_L$, that $\langle\theta_0 - \hat{\theta}\rangle^2$ converges almost surely to zero and that $\sqrt{L}\langle\theta_0 - \hat{\theta}\rangle^2$ is asymptotically normally distributed as $L\rightarrow \infty$.  However, $\hat{\rho}$ is not a consistent estimator of the amplitude $\rho_0$.  The asymptotic bias of $\hat{\rho}$ is small when the signal to noise ratio (SNR) is large, but the asymptotic bias is significant when the SNR is small.  %In Section~\ref{sec:gaussian-noise-case} we consider the special case when the noise is Gaussian.  In this case, our formula for the asymptotic variance of can be simplified.  
Section~\ref{sec:simulations} presents the results of Monte-Carlo simulations.  These simulations agree with the derived asymptotic properties.

@@

Before describing the statistical properties of the least squares estimator, we first require some properties of complex valued random variables.  
A complex random variable $W$ is said to be \emph{circularly symmetric} if its phase $\angle{W}$ is independent of its magnitude $\abs{W}$ and if the distribution of $\angle{W}$ is uniform on $[0,2\pi)$.  That is, if $Z \geq 0$ and $\Theta \in [0,2\pi)$ are real random variables such that $Ze^{j\Theta} = X$, then $\Theta$ is uniformly distributed on $[0,2\pi)$ and is independent of $Z$.  If the probability density function (pdf) of $Z$ is $f_Z(z)$, then the joint pdf of $\Theta$ and $Z$ is $f_{Z,\Theta}(z,\theta) = \frac{1}{2\pi}f_Z(z)$.
If $X$ is circularly symmetric, then for any real number $\phi$, the distribution of $X$ is the same as that of $e^{j\phi}X$.  %If $\expect\abs{X} = \expect Z$ is finite, then $X$ has zero mean because
% \begin{align*}
% \expect X &= \int_{0}^{2\pi} \int_{0}^\infty z e^{j\theta} f_{Z,\Theta}(z,\theta) dz d\theta \\
% &= \frac{1}{2\pi} \int_{0}^{2\pi} e^{j\theta} \int_{0}^\infty z f_Z(z) dz d\theta \\
% &= \frac{1}{2\pi}\expect Z \int_{0}^{2\pi} e^{j\theta} d\theta = 0.
% \end{align*}

% We will have particular use of complex random variables of the form $1 + X$ where $X$ is circularly symmetric.  Let $R \geq 0$ and $\Phi \in [0,2\pi)$ be real random variables satisfying $R e^{j\Phi} = 1 + X$.  The joint distribution of $R$ and $\Phi$ can be shown to be
% \[
% f(r,\phi) = \frac{r f_Z(\sqrt{r^2 - 2r\cos\phi + 1})}{2\pi\sqrt{r^2 - 2r\cos\phi + 1}}.
% \]

@@

\begin{theorem}\label{thm:normality} (Asymptotic normality)
Under the same conditions as Theorem~\ref{thm:consistency}, let $f(r,\phi)$ be the joint pdf of $R_1$ and $\Phi_1$, and let
\[
g(\phi) = \int_{0}^{\infty} r f(r,\phi) dr.
\]
Put $\hat{\lambda}_L = \sfracpart{\theta_0 - \hat{\theta}}$ and $\hat{m}_L = \hat{\rho} - \rho_0 G(0)$. %If $g(\phi)$ is continuous at $\phi = \tfrac{2\pi}{M}k+\tfrac{\pi}{M}$ for each $k = 0, 1, \dots M-1$, then 
Then the distribution of $(\sqrt{L}\hat{\lambda}_L, \sqrt{L}\hat{m}_L)$ converges to the bivariate normal with zero mean and covariance matrix
\[
\left( \begin{array}{cc} 
H^{-2} A & 0 \\
0 & \rho_0^2 B
\end{array} \right)
\]
as $L \rightarrow \infty$, where
\[
H = G(0) -  2 \sin(\tfrac{\pi}{M}) \sum_{k = 0}^{M-1} g(\tfrac{2\pi}{M}k + \tfrac{\pi}{M}),
\]
\[
A = \expect R_1^2\sin^2\fracpart{\Phi_1}, \;\;\; B = \expect R_1^2 \cos^2\fracpart{\Phi_1} - G^2(0). 
\]
\end{theorem}

@@

A key assumption in Theorem~\ref{thm:consistency} is that $G(x)$ is uniquely maximised at $x = 0$ for $x \in [-\tfrac{\pi}{M},\tfrac{\pi}{M})$.  %This assumption asserts that $G(x) \leq G(0)$ for all $x \in [-\pi, \pi)$ and that if $\{x_i\}$ is a sequence of numbers from $[-\pi,\pi)$ such that $G(x_i) \rightarrow G(0)$ as $i \rightarrow \infty$ then $x_i \rightarrow 0$ as $i \rightarrow \infty$.  
Although we will not prove it here, this assumption is not only sufficient, but also necessary, for if $G(x)$ is uniquely maximised at some $x \neq 0$ then $\sfracpart{\theta_0 - \hat{\theta}}_\pi \rightarrow x$ almost surely as $L\rightarrow\infty$, while if $G(x)$ is not uniquely maximised then $\sfracpart{\theta_0 - \hat{\theta}}_\pi$ will not converge.  One can check that the assumption holds when $w_1$ is circularly symmetric and normally distributed.  %We will not attempt to further classify those distributions for which the assumption holds here.
%The next proposition describes a class of circularly symmetric distributions for which $G(x)$ is uniquely maximised at $x = 0$.  The zero mean complex Gaussian distribution with independent real and imaginary parts is in this class.  We omit the proof due to space constraints.

@@

The theorems place conditions on $\sfracpart{\theta_0 - \hat{\theta}}$ rather than directly on $\theta_0 - \hat{\theta}$.  This makes practical sense, and to see why let $s_i' = e^{j2\pi k/M}s_i$ be $M$-PSK symbols obtained by rotating $s_1, \dots, s_L$ by $e^{j2\pi k/M}$ for some integer $k$.  Then
\[
a s_i =  \rho e^{j\theta}s_i = \rho e^{j(\theta - 2\pi k/M)} s_i',
\]
and so, if $a = \hat{\rho}e^{j\hat{\theta}}$ minimises~\eqref{eq:hatadef}, then so does $\hat{\rho} e^{j(\hat{\theta} - 2\pi k/M)}$.  Thus, $\theta_0 - \hat{\theta}$ should be attributed the same error as $\theta_0 - \hat{\theta} - \tfrac{2\pi}{M}k$.  That is, the error should be computed `modulo $\tfrac{2\pi}{M}$'.  It is for this reason that \emph{differential encoding} is often used with noncoherent $M$-PSK  detectors.

### 2012-10-26 paper.tex
Added more of Barry's edits.

The existing literature~\cite{Mackenthun1994,Cowley_ref_sym_carr_1998,ViterbiViterbi_phase_est_1983,Sweldens2001,Wilson1989,Makrakis1990,Liu1991} mostly considers what is called \emph{noncoherent detection} where no pilot symbols exist ($P = \emptyset$ where $\emptyset$ is the empty set).  In the nocoherent setting \emph{differential encoding} is often used, and for this reason the estimation problem has been called \emph{multiple symbol differential detection}.  A popular approach is the so called \emph{non-data aided}, sometimes also called \emph{non-decision directed}, estimator based on the paper of Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983}.  The idea is to `strip' the modulation from the recieved signal by taking $y_i / \abs{y_i}$ to the power of $M$.  A function $F: \reals \mapsto \reals$ is chosen and the estimator of the carrier phase $\theta_0$ is taken to be $\tfrac{1}{M}\angle{A}$ where $\angle$ denotes the complex argument and
\[
A = \frac{1}{L}\sum_{i \in P \cup D} F(\abs{y_i}) \big(\tfrac{y_i}{\abs{y_i}}\big)^M.
\]
Various choices for $F$ are suggested in~\cite{ViterbiViterbi_phase_est_1983} and a statistical analysis is presented.  A caveat of this estimator is that it is not entirely obvious how pilot symbols should be included. Techniques are available but they are somewhat adhoc~\cite{Cowley_ref_sym_carr_1998}.  This problem does not occur with the least square estimator.

@@

In the literature it has been common to assume that the data symbols $\{d_i, i \in D\}$ are of primary interest and that the complex amplitude $a_0$ is a nuisance parameter.  The metric of performance is correspondingly the \emph{symbol error rate}, or \emph{bit error rate}.  While estimating the symbols (or more precisely the transmitted bits) is ultimately the goal, we take the opposite point of view here.  Our aim is to estimate $a_0$, and we treat the unknown data symbols as nuisance parameters.  This is motivated by the fact that in many modern communication systems the data symbols are \emph{coded}.  For this reason raw symbol error rate is not of interest at this stage.  Instead, we desire an accurate estimator $\hat{a}$ of $a_0$, so that the compensated recieved symbols $\hat{a}^{-1}y_i$ can be accurately modelled using an additive noise channel.  The additive noise channel is a common assumption for subsequent reciever operations, such as decoding.  The estimator $\hat{a}$ is also used in the computation of decoder metrics for modern decoders, and for interference cancellation in multiuser systems.  Consequently, our metric of performance will not be symbol or bit error rate, but $\abs{a_0 - \hat{a}}^2$. It will be informative to consider the carrier phase and amplitude estimators separately, that is, if $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ where $\hat{\rho}$ is a positive real number, then we consider $\langle\theta_0 - \hat{\theta}\rangle_\pi^2$ and $(\rho_0 - \hat{\rho})^2$.  The function $\fracpart{\cdot}_\pi$ denotes its argument taken `modulo $2\pi$' into the interval $[-\pi, \pi)$.  It will become apparent why $\langle\theta_0 - \hat{\theta}\rangle_\pi^2$ rather than $(\theta_0 - \hat{\theta})^2$ is the appropriate measure of error for the phase parameter.

@@

In this section we derive Mackentun's algorithm to compute the least squares estimator of the carrier phase and amplitude~\cite{Mackenthun1994}.  Mackenthun specifically considered the noncoherent setting, so we modify the algorithm to include the pilot symbols.  For the purpose of anaylsing computational complexity, we will assume that the number of data symbols $\abs{D}$ is proportional to the total number of symbols $L$, so that, for example, $O(L) = O(\abs{D})$.  In this case Mackentun's algorithm requires $O(L \log L)$ arithmetic operations.  This complexity arises from the need to sort a list of $\abs{D}$ elements.  
%We use order notation in the standard way, that is, for functions $h$ and $g$, we write $h(N) = O(g(N))$ to mean that there exists a constant $K > 0$ and a finite $N_0$ such that $h(N) \leq K g(N)$ for all $N > N_0$.

@@

The theorems place conditions on $\sfracpart{\hat{\theta} - \theta_0}_\pi$ rather than directly on $\hat{\theta} - \theta_0$.  This makes sense because the phases $\theta_0$ and $\theta_0 + 2\pi k$ are equivalent for any integer $k$. So, for example, we expect the phases $0.49\pi$ and $-0.49\pi$ to be close together, the difference between them being $\vert\sfracpart{-0.49\pi - 0.49\pi}_\pi\vert = 0.02\pi$, and not $\vert -0.49\pi - 0.49\pi\vert = 0.98\pi$.

@@

Finally, when $k < 0$ or $k > M$, 
\[
\int_0^{2\pi} v_k(\lambda, \phi) g(\phi) d\phi = 0.
\]
So the infinite sum in~\eqref{eq:Winfsum} can be written as the finite sum,
\begin{align*}
W(\lambda) &= \sum_{k=0}^{M} \int_{0}^{2\pi} v_k(\lambda, \phi) g(\phi) d\phi \\
&= \sum_{k=0}^{M-1} \lambda c_k(\lambda) + \sum_{k=0}^{M-1} \lambda a_k(\lambda) + \sum_{k=0}^{M} \lambda b_k(\lambda) \\
&= \lambda \big( C(\lambda)  + A(\lambda) + B(\lambda) \big),
\end{align*}
where 
\[
A(\lambda) = \sum_{k=0}^{M-1} a_k(\lambda), \qquad B(\lambda) = \sum_{k=0}^{M} b_k(\lambda),  
\]
\[
C(\lambda) = \sum_{k=0}^{M-1} c_k(\lambda).
\]
Both $A(\lambda)$ and $B(\lambda)$ converge to zero as $\lambda \rightarrow 0$, whilst $C(\lambda)$ converges to $H_1$ as $\lambda$ converges to zero.  The lemma follows after puting $\gamma(\lambda) = A(\lambda) + B(\lambda) + C(\lambda) - H_1$.
\end{IEEEproof}

\begin{lemma}\label{lem:empiricprocc} Let $Q_L(\lambda) = \expect R_L(\lambda) - \expect R_L(0)$ where the function $R_L$ is defined in~\eqref{eq:RLdef}.  Then,
\[
\sqrt{L}\big( R_L(\hat{\lambda}_L) - Q_L(\hat{\lambda}_L) \big) = o_P(1) + \sqrt{L} R_L(0).
\]
\end{lemma}
\begin{IEEEproof}
Write
\[
\sqrt{L}\big( R_L(\hat{\lambda}_L) - Q_L(\hat{\lambda}_L) \big) = W_L(\hat{\lambda}_L) + \sqrt{L} R_L(0)
\]
where
\begin{equation}\label{eq:WLdef}
W_L(\lambda) = \sqrt{L}\big( R_L(\lambda) - Q_L(\lambda) - R_L(0) \big)
\end{equation}
is what is called an \emph{empirical process} indexed by $\lambda$~\cite{Pollard_asymp_empi_proc_1989,Pollard_new_ways_clts_1986,van2009empirical,Pollard_conv_stat_proc_1984}.  Techniques from this literature can be used to show that
%Lemma~\ref{lem:ZYLempiricproc} 
%in Section~\ref{sec:an-empirical-process} 
%shows that 
for any $\delta > 0$ and $\nu > 0$, there exists $\epsilon > 0$ such that
\[
\Pr\left\{ \sup_{\sabs{\lambda}<\epsilon}\sabs{W_L(\lambda)} > \delta  \right\} < \nu
\]
for all positive integers $L$.  This type of result is typically called \emph{tightness} or \emph{asymptotic continuity}~\cite{Pollard_asymp_empi_proc_1989,van2009empirical,Billingsley1999_convergence_of_probability_measures}.  We ommit the proof which follows in a straightforward, but lengthy manner using an argument called \emph{symmetrisation} followed by an argument called \emph{chaining}~\cite{Pollard_asymp_empi_proc_1989,van2009empirical}.

>>>>

The existing literature~\cite{Mackenthun1994,Cowley_ref_sym_carr_1998,ViterbiViterbi_phase_est_1983,Sweldens2001,Wilson1989,Makrakis1990,Liu1991} mostly considers what is called \emph{noncoherent detection} where no pilot symbols exist ($P = \emptyset$ where $\emptyset$ is the empty set).  In the noncoherent setting \emph{differential encoding} is often used, and for this reason the estimation problem has been called \emph{multiple symbol differential detection}.  A popular approach is the so called \emph{non-data aided}, sometimes also called \emph{non-decision directed}, estimator based on the paper of Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983}.  The idea is to `strip' the modulation from the received signal by taking $y_i / \abs{y_i}$ to the power of $M$.  A function $F: \reals \mapsto \reals$ is chosen and the estimator of the carrier phase $\theta_0$ is taken to be $\tfrac{1}{M}\angle{A}$ where $\angle$ denotes the complex argument and
\[
A = \frac{1}{L}\sum_{i \in P \cup D} F(\abs{y_i}) \big(\tfrac{y_i}{\abs{y_i}}\big)^M.
\]
Various choices for $F$ are suggested in~\cite{ViterbiViterbi_phase_est_1983} and a statistical analysis is presented.  A caveat of this estimator is that it is not entirely obvious how pilot symbols should be included. Techniques are available but they are somewhat adhoc~\cite{Cowley_ref_sym_carr_1998}.  This problem does not occur with the least square estimator.

@@

In the literature it has been common to assume that the data symbols $\{d_i, i \in D\}$ are of primary interest and that the complex amplitude $a_0$ is a nuisance parameter.  The metric of performance is correspondingly the \emph{symbol error rate}, or \emph{bit error rate}.  While estimating the symbols (or more precisely the transmitted bits) is ultimately the goal, we take the opposite point of view here.  Our aim is to estimate $a_0$, and we treat the unknown data symbols as nuisance parameters.  This is motivated by the fact that in many modern communication systems the data symbols are \emph{coded}.  For this reason raw symbol error rate is not of interest at this stage.  Instead, we desire an accurate estimator $\hat{a}$ of $a_0$, so that the compensated received symbols $\hat{a}^{-1}y_i$ can be accurately modelled using an additive noise channel.  The additive noise channel is a common assumption for subsequent receiver operations, such as decoding.  The estimator $\hat{a}$ is also used in the computation of decoder metrics for modern decoders, and for interference cancellation in multiuser systems.  Consequently, our metric of performance will not be symbol or bit error rate, but $\abs{a_0 - \hat{a}}^2$. It will be informative to consider the carrier phase and amplitude estimators separately, that is, if $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ where $\hat{\rho}$ is a positive real number, then we consider $\langle\theta_0 - \hat{\theta}\rangle_\pi^2$ and $(\rho_0 - \hat{\rho})^2$.  The function $\fracpart{\cdot}_\pi$ denotes its argument taken `modulo $2\pi$' into the interval $[-\pi, \pi)$.  It will become apparent why $\langle\theta_0 - \hat{\theta}\rangle_\pi^2$ rather than $(\theta_0 - \hat{\theta})^2$ is the appropriate measure of error for the phase parameter.

@@

In this section we derive Mackentun's algorithm to compute the least squares estimator of the carrier phase and amplitude~\cite{Mackenthun1994}.  Mackenthun specifically considered the noncoherent setting, so we modify the algorithm to include the pilot symbols.  For the purpose of analysing computational complexity, we will assume that the number of data symbols $\abs{D}$ is proportional to the total number of symbols $L$, so that, for example, $O(L) = O(\abs{D})$.  In this case Mackentun's algorithm requires $O(L \log L)$ arithmetic operations.  This complexity arises from the need to sort a list of $\abs{D}$ elements.  
%We use order notation in the standard way, that is, for functions $h$ and $g$, we write $h(N) = O(g(N))$ to mean that there exists a constant $K > 0$ and a finite $N_0$ such that $h(N) \leq K g(N)$ for all $N > N_0$.

@@

The theorems place conditions on $\sfracpart{\theta_0 - \hat{\theta}}_\pi$ rather than directly on $\theta_0 - \hat{\theta}$.  This makes sense because the phases $\theta_0$ and $\theta_0 + 2\pi k$ are equivalent for any integer $k$. So, for example, we expect the phases $0.49\pi$ and $-0.49\pi$ to be close together, the difference between them being $\vert\sfracpart{-0.49\pi - 0.49\pi}_\pi\vert = 0.02\pi$, and not $\vert -0.49\pi - 0.49\pi\vert = 0.98\pi$.

@@

Finally, when $k < 0$ or $k > M$, 
\[
\int_0^{2\pi} v_k(\lambda, \phi) g(\phi) d\phi = 0.
\]
So the infinite sum in~\eqref{eq:Winfsum} can be written as the finite sum,
\begin{align*}
W(\lambda) &= \sum_{k=0}^{M} \int_{0}^{2\pi} v_k(\lambda, \phi) g(\phi) d\phi \\
&= \sum_{k=0}^{M-1} \lambda c_k(\lambda) + \sum_{k=0}^{M-1} \lambda a_k(\lambda) + \sum_{k=0}^{M} \lambda b_k(\lambda) \\
&= \lambda \big( C(\lambda)  + A(\lambda) + B(\lambda) \big),
\end{align*}
where 
\[
A(\lambda) = \sum_{k=0}^{M-1} a_k(\lambda), \qquad B(\lambda) = \sum_{k=0}^{M} b_k(\lambda),  
\]
\[
C(\lambda) = \sum_{k=0}^{M-1} c_k(\lambda).
\]
Both $A(\lambda)$ and $B(\lambda)$ converge to zero as $\lambda \rightarrow 0$, whilst $C(\lambda)$ converges to $H_1$ as $\lambda$ converges to zero.  The lemma follows after putting $\gamma(\lambda) = A(\lambda) + B(\lambda) + C(\lambda) - H_1$.
\end{IEEEproof}

\begin{lemma}\label{lem:empiricprocc} Let $Q_L(\lambda) = \expect R_L(\lambda) - \expect R_L(0)$ where the function $R_L$ is defined in~\eqref{eq:RLdef}.  Then,
\[
\sqrt{L}\big( R_L(\hat{\lambda}_L) - Q_L(\hat{\lambda}_L) \big) = o_P(1) + \sqrt{L} R_L(0).
\]
\end{lemma}
\begin{IEEEproof}
Write
\[
\sqrt{L}\big( R_L(\hat{\lambda}_L) - Q_L(\hat{\lambda}_L) \big) = W_L(\hat{\lambda}_L) + \sqrt{L} R_L(0)
\]
where
\begin{equation}\label{eq:WLdef}
W_L(\lambda) = \sqrt{L}\big( R_L(\lambda) - Q_L(\lambda) - R_L(0) \big)
\end{equation}
is what is called an \emph{empirical process} indexed by $\lambda$~\cite{Pollard_asymp_empi_proc_1989,Pollard_new_ways_clts_1986,van2009empirical,Pollard_conv_stat_proc_1984}.  Techniques from this literature can be used to show that
%Lemma~\ref{lem:ZYLempiricproc} 
%in Section~\ref{sec:an-empirical-process} 
%shows that 
for any $\delta > 0$ and $\nu > 0$, there exists $\epsilon > 0$ such that
\[
\Pr\left\{ \sup_{\sabs{\lambda}<\epsilon}\sabs{W_L(\lambda)} > \delta  \right\} < \nu
\]
for all positive integers $L$.  This type of result is typically called \emph{tightness} or \emph{asymptotic continuity}~\cite{Pollard_asymp_empi_proc_1989,van2009empirical,Billingsley1999_convergence_of_probability_measures}.  We omit the proof which follows in a straightforward, but lengthy manner using an argument called \emph{symmetrisation} followed by an argument called \emph{chaining}~\cite{Pollard_asymp_empi_proc_1989,van2009empirical}.
